---
title: "Wavelet Screaming on methyaltion"
author: "William Denault"
date: "19/02/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Document/Serieux/Travail/Data_analysis_and_papers/Fast functional association analysis based on wavelets/additionnal_simulation_AOAS")
```

## Preparing the data


```{r cars}
load("simulated_data.Rdata")
pheno <- as.numeric(colon.state)-1

```

### Definin slice 

1.89 percent missigness
```{r }
thresh <- 500
tl <- split(temp.pos , cumsum(c(1, diff(temp.pos) > thresh)  ) ) #Remove case with too long distance

min(lengths(tl))
#Level of analysis 4
tl <-  tl[-which(lengths(tl)==16)]
tl <-  tl[-which(lengths(tl)==15)]

sum(lengths(tl))/length( temp.pos)#percent missigness
hist( lengths(tl),nclass = 100)

```


## Running Wavelet Screaming slice by slice


```{r message=FALSE,  echo=FALSE, results="markup"}
library(WaveletScreaming)
res <- list()
for ( i in 1 :length(tl))
{
  reg_mat <- sim_methy_all[which( temp.pos  %in% tl[[i]]),]  
  res[[i]] <- Wavelet_screaming( Y=pheno,
                                 loci=reg_mat,
                                 lev_res=4,coeftype = "c",
                                 bp = temp.pos[which( temp.pos  %in% tl[[i]])],
                                 base_shrink = 1/26,
                                 sigma_b =200# 1/sqrt(length(pheno))
  )
  
  
}

```


```{r echo=TRUE}
res <- do.call( rbind, res)
res[1:10,]

```

## Simulate null distribution
```{r ,message=FALSE,}
set.seed(1)
Sim <- Simu_null_emp(res=res,coeftype="c",
                       lev_res = 4,
                       base_shrink = 1/26,
                       size=100000)
```


## Performing the Box-Cox transform
```{r ,message=FALSE,echo=TRUE}
library(EnvStats)

bc_lambda <-  boxcox(-Sim[,1], optimize = TRUE)
Sim[,1]  <-  - (((-Sim[,1])^bc_lambda$lambda) -1  )/bc_lambda$lambda
res[,1]  <-  - (((-res[,1])^bc_lambda$lambda) -1  )/bc_lambda$lambda
dat <- data.frame(dens = c(c(res[,1]),c(Sim[,1]))
                  , lines = c(rep("obs", length(c(res[,1]))), rep( "sim", length(c(Sim[,1]))) ))
#Plot.
ggplot(dat, aes(x = dens, fill = lines)) + geom_density(alpha = 0.5)+ggtitle("Test stat density")

```


## Computing the p-values

```{r,message=FALSE }
lambda <- Search_lambda(Sim,plot=TRUE)

Th <- Sim[,c("L_h")]+lambda*Sim[,c("min_ph_pv")]
muv <- median(Th,na.rm = TRUE)
sdv <- mad(Th,na.rm = TRUE)
####################################
##Test Value of the loci to be tested
####################################
th <-  res[,1]+lambda*res[,2]

dat <- data.frame(dens = c(c(th),c(Th))
                  , lines = c(rep("obs", length(c(th))), rep( "sim", length(c(Th))) ))
#Plot.
ggplot(dat, aes(x = dens, fill = lines)) + geom_density(alpha = 0.5)+ggtitle("Test stat density")
```




```{r,message=FALSE }

pv <- 1-pnorm(th,mean=muv,sd=sdv)
hist(1-pnorm(th,mean=muv,sd=sdv),nclass=100)
hist(log10(1-pnorm(th,mean=muv,sd=sdv)))
```



## Loading data from Irizzary et al 2009 Nature Genetics
```{r,message=FALSE }
library(readxl)
tt <- read_excel("D:/Document/Serieux/Travail/Data_analysis_and_papers/Fast functional association analysis based on wavelets/additionnal_simulation_AOAS/41588_2009_BFng298_MOESM18_ESM.xls")
sub <- tt[which(tt[,2]=="chr3"),]
sub <- as.data.frame(sub
)

lstemp <- list()

for (i in 1: dim(sub)[1])
{
  
  lstemp[[i]] <- temp.pos[ which( (sub$start[i]-1)  < temp.pos & temp.pos < (sub$end[i] +1 ))]
  
}

sum( lengths(lstemp)) # the list referred by Lee and Morris

pos_to_detect <-  do.call( c,lstemp)#List of the CpG to detect


```



## Performing a slicing 

Ensure same slicing  than before

```{r }


thresh <- 500

tl <- split(temp.pos , cumsum(c(1, diff(temp.pos) > thresh)  ) ) #Remove case with too long distance
tl <-  tl[-which(lengths(tl) <17)]#Keep only long enough sequence of CPG

pos_to_detect <- pos_to_detect [which(pos_to_detect %in% do.call( c , tl))] #filtering out the CpG remove when slicing

```


## Power the if a region contains more than one CpG that is differentially methylated then it has to be deteceted
```{r }
#N° of CpG per regions
n_CpG_per_region <- list()




for ( i in 1:length(tl))
{
  n_CpG_per_region[[i]] <- pos_to_detect[which(pos_to_detect %in% tl[[i]])]

}


nCpG_reg <- lengths(n_CpG_per_region)
indx_to_detect <- which( nCpG_reg >0)
indx_to_detect


```


When using p-values 10^-5
```{r}
indx_WS <-  which(pv < 10^(-5))
length(indx_WS)
sum(  indx_WS %in% indx_to_detect)

which(indx_WS %in% indx_to_detect)
```

When using p-values 10^-6
```{r}
indx_WS <-  which(pv < 10^(-6))
length(indx_WS)
sum(  indx_WS %in% indx_to_detect)

which(indx_WS %in% indx_to_detect)
```


When using FDR

```{r}

fdr_regions <- p.adjust(pv, method ="BH")
indx_WS <-  which(fdr_regions  <0.05)
length(indx_WS)
which(indx_WS %in% indx_to_detect)
sum(  indx_WS %in% indx_to_detect)

```

```{r}
fdr_regions <- p.adjust(pv, method ="BH")
indx_WS <-  which(fdr_regions  <0.01)
length(indx_WS)
which(indx_WS %in% indx_to_detect)
sum(  indx_WS %in% indx_to_detect)

```

