---
title: "Wavelet Screening applied to DNA methylation"
author: "William R.P. Denault et al."
date: "12/01/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Preparing the data

Loading the data and recoding the phenotype from factor to numeric.


```{r cars}

load("simulated_data.Rdata")
colon.state[1:4]
pheno <- as.numeric(colon.state)-1
pheno[1:4]

```

The data are a compound of two elements:
1) The base pair position of each CpG,
2) The individual CpG level (M value)
```{r}
temp.pos[1:10]#the base pair position of each CpG
length(temp.pos)#total number of CpGs
dim(sim_methy_all)#lines= CpG level, column individual
sim_methy_all[1:10,]
```





### Defining slice 
Here, we divide the data into regions that have at least 17 CpGs, separated by a maximum distance of 500 bp. This results in a missingness rate of 1.89%.
```{r }

thresh <- 500
tl <- split(temp.pos , cumsum(c(1, diff(temp.pos) > thresh)  ) ) #Remove cases that are too far apart

#Level of analysis 4
tl <-  tl[-which(lengths(tl)< 17)]#List of the positions of CpGs that we will analyze later
sum(lengths(tl))/length( temp.pos)#percent missingness
hist( lengths(tl),nclass = 100, main="Histogram of the number of CpGs per region")

```


## Running Wavelet Screening region by region


In this section, we run Wavelet Screening on each region separately. This takes about 3 to 4 minutes on a single CPU. 
```{r message=FALSE}

library(WaveletScreening)
tt <- proc.time()
res <- list()
for ( i in 1 :length(tl))
{
  reg_mat <- sim_methy_all[which( temp.pos  %in% tl[[i]]),] #Select CpGs in region i
  bp <- temp.pos[which( temp.pos  %in% tl[[i]])]#Correspond to the base position of the CpG in region i
  res[[i]] <- Wavelet_screening( Y=pheno,
                                 loci=reg_mat,
                                 lev_res=4,
                                 coeftype = "c",
                                 bp = bp,
                                 base_shrink = 1/26,
                                 sigma_b =200,
                                 verbose = FALSE
  )


}


proc.time()  -tt #total run time
```

We concatenate the results of Wavelet Screening into one data frame. 
```{r echo=TRUE}
res <- do.call( rbind, res)
res[1:10,]
```

## Simulation of the null distribution

To assess the significance of each region, we simulate the null distribution of the test statistic. Below we display how to simulate this distribution.
```{r ,message=FALSE,}
set.seed(1)
#Simulation function using the same parameter as used in the Wavelet_screening function
tt <- proc.time()
Sim <- Simu_null_emp(res=res,
                     coeftype="c",
                     lev_res = 4,
                     base_shrink = 1/26,
                     size=100000#Number of simulations required
)
proc.time()  -tt #total running time
plot( Sim[,1], Sim[,2])
```


## Performing the Box-Cox transform

As explained in the article, the distribution of L~h~ might not be Gaussian when analyzing data using low depth. Below, we display how to perform a Box-Cox transform on the simulations of the test statistic under the H~0.

Here, we have an additionnal issue: the observed test statistics contains some positive and some negative values. However the Box-Cox transform requires only positive values. We therefore shift the distribution and then apply the Box-Cox transform.
```{r ,message=FALSE,echo=TRUE}
library(EnvStats)
shifting  <-  max( 0,1.0001*max( Sim[,1],res[,1]))
bc_lambda <-  boxcox(-Sim[,1] + shifting, optimize = TRUE)
Sim[,1]   <-  - ( ( (-Sim[,1] + shifting)^bc_lambda$lambda) -1  )/bc_lambda$lambda
res[,1]   <-  - ( ( (-res[,1] + shifting)^bc_lambda$lambda) -1  )/bc_lambda$lambda

```


## Computing the p-values

When the Box-Cox transform is done, we can then compute the p-value of each region as follows.
```{r,message=FALSE }
lambda <- Search_lambda(Sim,plot=TRUE)#optimizing the lambda* value

Th <- Sim[,c("L_h")]+lambda*Sim[,c("min_ph_pv")] #Computing the null distribution of the test statistic 
muv <- median(Th,na.rm = TRUE) #estimating the parameter of the null distribution
sdv <- mad(Th,na.rm = TRUE)
####################################
##Test value of the loci to be tested
####################################
th <-  res[,1]+lambda*res[,2]#Computing the test statistic

##########################
##Plot of two distributions
##########################
dat <- data.frame(dens = c(c(th),c(Th[1:2000]))#To have the same size on the plot
                  , lines = c(rep("obs", length(c(th))), rep( "sim", length(c(Th[1:2000]))) ))

ggplot(dat, aes(x = dens, fill = lines)) +
  geom_density(alpha = 0.5)+
  xlim(c(-3,3))+#Some of the associated regions have test statistics very far from the null; this results in poor visualization if they are not excluded.
  ggtitle("Test stat density")
```

Here, we see the null distribution of the test statistic, and the associated regions on the right side of the plot.

```{r,message=FALSE }
#Computing the p-value
pv <- 1-pnorm(th,mean=muv,sd=sdv)
hist(1-pnorm(th,mean=muv,sd=sdv),nclass=100, main="Histogram of the pvalues")
hist(log10(1-pnorm(th,mean=muv,sd=sdv)),nclass=100,main ="Histogram of the Log10 pvalues")
```



## Loading data from Irizzary et al. 2009  
The data can be found at https://www.nature.com/articles/ng.298 as supplementary material.

```{r,message=FALSE }
library(readxl)
tt <- read_excel("41588_2009_BFng298_MOESM18_ESM.xls")
sub <- tt[which(tt[,2]=="chr3"),]#Only selecting Chromosome 3
sub <- as.data.frame(sub)
head(sub) #data frame with the start and end positions of each DMR we are trying to detect.

lstemp <- list() #

for (i in 1: dim(sub)[1])
{
  
  lstemp[[i]] <- temp.pos[ which( (sub$start[i]-1)  < temp.pos & temp.pos < (sub$end[i] +1 ))]
  
}

sum( lengths(lstemp)) # number of CpGs to detect
pos_to_detect <-  do.call( c,lstemp)#List of the CpGs to detect


```



## Performing a slicing 

Ensure the same slicing as before.

```{r }
thresh <- 500

tl <- split(temp.pos , cumsum(c(1, diff(temp.pos) > thresh)  ) ) #Remove cases that are too far apart
tl <-  tl[-which(lengths(tl) <17)]#Keep only long-enough sequences of CPGs

pos_to_detect <- pos_to_detect [which(pos_to_detect %in% do.call( c , tl))] #filtering out the CpGs when slicing

```


## Power

If a region contains more than one CpG that is differentially methylated, then it has to be detected.
```{r }
#Number of CpGs per regions
n_CpG_per_region <- list()

for ( i in 1:length(tl))
{
  n_CpG_per_region[[i]] <- pos_to_detect[which(pos_to_detect %in% tl[[i]])]

}


nCpG_reg <- lengths(n_CpG_per_region)
indx_to_detect <- which( nCpG_reg >0)#List of the regions to detect
indx_to_detect


```



Regions detected
```{r}
type <- c("pv", "pv", "fdr", "fdr")
threshold <- c("e-5", "e-6", "5%", "1%")
detected_reg <- c()
tru_reg <- c()
#Regions detected

#When using p-values 10^-5
indx_WS <-  which(pv < 10^(-5))
detected_reg <- c( detected_reg ,length(indx_WS))
tru_reg <- c( tru_reg , length(which(indx_WS %in% indx_to_detect) ) )

#When using p-values 10^-6
indx_WS <-  which(pv < 10^(-6))
detected_reg <- c( detected_reg ,length(indx_WS))
tru_reg <- c( tru_reg , length(which(indx_WS %in% indx_to_detect) ) )
fdr_regions <- p.adjust(pv, method ="BH")

#When using FDR at 5%
indx_WS <-  which(fdr_regions  <0.05)
detected_reg <- c( detected_reg ,length(indx_WS))
tru_reg <- c( tru_reg , length(which(indx_WS %in% indx_to_detect) ) )

#When using FDR at 1%
indx_WS <-  which(fdr_regions  <0.01)
detected_reg <- c( detected_reg ,length(indx_WS))
tru_reg <- c( tru_reg , length(which(indx_WS %in% indx_to_detect) ) )

#Summary 
df_out <- data.frame(type=type, threshold = threshold, nb_detected_region= detected_reg, nb_true_regions= tru_reg )
```
```{r}
df_out
```

df_out summarizes the number of regions detected using different criteria and the number of truly associated regions.
